---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r include=FALSE} 
# Global Options: 
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, fig.path = 'figure/') 
```
## Loading and preprocessing the data
```{r}
# Get he current working directory
wd <- getwd();
# Extract the data set if the data set folder does not exists
if(!dir.exists(file.path(wd, 'activity'))){
  zip <- paste(wd, 'activity.zip', sep = '/');
  outDir <- paste(wd, 'activity', sep = '/');
  unzip(zip,exdir=outDir)
}

# Read the data set
ds <- read.csv2('activity/activity.csv', sep = ',')
```


## What is mean total number of steps taken per day?
```{r}
# Find all Na rows in ds$steps
x <- is.na(ds$steps)
# Remove the NA rows and store the new data.frame as df
df <- ds[!x,]
# Load dplyr for the summary stats
library(dplyr)
# Use piping to create a new data.frame with all the daily steps grouped by say
sums <- df %>% group_by(date) %>%
        summarise(total.steps = sum(steps), mean.steps = mean(steps),
                  median.steps = median(steps))

library(ggplot2)

# Find the quantiles
cols <- quantile(sums$total.steps)

# Create quantile groups
sums$group <- ifelse(sums$total.steps <= cols[1], 'a',
                     ifelse(sums$total.steps <= cols[2], 'b',
                            ifelse(sums$total.steps <= cols[4], 'c', 
                                   ifelse(sums$total.steps <= cols[5], 
                                          'd', 'e'))))
# Set a variable to store the alpha value
a = 0.65
# Create a histogram of the total steps per day, coloured by quantile group
hg <- ggplot(sums, aes(x = total.steps)) + 
  geom_histogram(binwidth = 450, data = subset(sums, group == 'a'), alpha = a,
                 aes(fill = group)) +
  geom_histogram(binwidth = 450, data = subset(sums, group == 'b'), alpha = a,
                 aes(fill = group))+
  geom_histogram(binwidth = 450, data = subset(sums, group == 'c'), alpha = a,
                 aes(fill = group))+
  geom_histogram(binwidth = 450, data = subset(sums, group == 'd'), alpha = a,
                 aes(fill = group)) +
  geom_histogram(binwidth = 450, data = subset(sums, group == 'e'), alpha = a,
                 aes(fill = group)) +
  ggtitle('Total Steps Per Day') + xlab('Total Steps Taken') + ylab('Density')+
  scale_fill_manual(name = 'group',
      values = c('red','orange', 'yellow', 'lightblue',
                               'aquamarine'),
                    labels = c('First quartile', 'Second Quartile', 
                               'Mid-range', 'Third Quarile', 'Fourth quartile'))
# now we save the file
ggsave(filename = 'histogram.png', plot = hg, device = 'png', path = 'figure/',
       width = 20, height = 15, units = 'cm', dpi = 320)
```
<img src = 'histogram.png' />
```{r}
# Mean steps per day
print('Mean steps a day: ')
print(sums[,c('date', 'mean.steps')])
```
```{r}
# Median steps per day
print('Median steps taken per day')
print(sums[c('date', 'median.steps')])
```

## What is the average daily activity pattern?
```{r}
# Create a new data.frame for the mean steps per 5 minute interval
pattern.df <- df %>% group_by(interval) %>%
  summarise(average.steps = mean(steps))
#Plot the data.frame, with an added horizontal line for the overall mean
hg <- ggplot(data = pattern.df, aes(x = interval, y = average.steps)) + geom_line(col = 'brown') +
  xlab('5 - Minute Interval') + ylab('Average Daily Steps') + 
  ggtitle('Daily Activity Pattern') + 
  geom_hline(yintercept =  mean(df$steps), show.legend = TRUE, col = 'green')
# Now we save the png
ggsave(filename = 'activity pattern.png', plot = hg, device = 'png', path = 'figure/',
       width = 20, height = 15, units = 'cm', dpi = 320)

  
```
<img src = 'activity pattern.png' />
```{r}
print('The inerval with the highest mean number of steps is: ')
print(pattern.df[max(pattern.df$average.steps) == pattern.df$average.steps,
                 "interval"])
```


## Imputing missing values
```{r}
# Coun the number of rows with Na values
total.nas <- sum((is.na(ds$steps) | is.na(ds$date) | is.na(ds$interval)))
print('The total number of rows wih an NA value is: ')
print(total.nas)
```
```{r}
# Using the mean value o impute all missing rows:
full.ds <- ds
# Fill in the missing data with the mean value. all other columns have no Nas
# So we only impute u=into steps.
full.ds[x, 'steps'] <- mean(df$steps)
#  Making he daily satistics
full.sums <- full.ds %>% group_by(date) %>% 
  summarise(total.steps = sum(steps), mean.steps = mean(steps),
            median.steps = median(steps))
# Make the hisogram of total daily steps:
# First calculate the quantiles
cols <- quantile(full.sums$total.steps)
# Next group the data.frame by quantile range 
full.sums$group <- ifelse(full.sums$total.steps <= cols[1], 'a',
                     ifelse(full.sums$total.steps <= cols[2], 'b',
                            ifelse(full.sums$total.steps <= cols[4], 'c', 
                                   ifelse(full.sums$total.steps <= cols[5], 
                                         'd', 'e'))))
# Set the alpha value
a = 0.65

# Plot the dataset
hg <- ggplot(full.sums, aes(x = total.steps)) + 
  geom_histogram(binwidth = 450, data = subset(full.sums, group == 'a'),
                 alpha = a, aes(fill = group)) +
  geom_histogram(binwidth = 450, data = subset(full.sums, group == 'b'),
                 alpha = a, aes(fill = group))+
  geom_histogram(binwidth = 450, data = subset(full.sums, group == 'c'),
                 alpha = a,aes(fill = group))+
  geom_histogram(binwidth = 450, data = subset(full.sums, group == 'd'),
                 alpha = a, aes(fill = group)) +
  geom_histogram(binwidth = 450, data = subset(full.sums, group == 'e'),
                 alpha = a, aes(fill = group)) +
  ggtitle('Total Steps Per Day') + xlab('Total Steps Taken') + ylab('Density')+
  scale_fill_manual(name = 'group',
      values = c('red','orange', 'yellow', 'pink',
                               'aquamarine'),
                    labels = c('First quartile', 'Second Quartile', 
                               'Mid-range', 'Third Quarile', 'Fourth quartile'))
# Save the png
ggsave(filename = 'imputed histogram.png', plot = hg, device = 'png', path = 'figure/',
       width = 20, height = 15, units = 'cm', dpi = 320)
```
<img src = 'figure/imputed histogram.png' />
```{r}
print('The mean number of steps taken per day for this dataset is: ')
print(full.sums$mean.steps)
```
```{r}
print('The median number of steps taken per day for this dataset is: ')
print(full.sums$median.steps)
```
As seen, imputing missing data changes the estimates as follows: 
* The mean number of steps per day is now esimated to be higher
* The median number of steps per day is non zero on the imputed rows
* Imputing missing values increases the estimated total number of steps per day
## Are there differences in activity patterns between weekdays and weekends?
```{r}
# Get the dates, turn them into a Date type, then turn them into week day names
days <- weekdays.Date(as.Date.factor(full.ds$date))
# Next we set them to 'weekend', or 'weekday' depending on the day type
days <- c(ifelse(days == 'Sunday' | days == 'Saturday', 'Weekend', 'Weekday'))
# Then we conver them into a factor and store it in the data.frame
full.ds$day.type <- as.factor(days)
# Summarise the data.frame
plot.df <- full.ds %>% group_by(interval, day.type) %>%
  summarise(mean.steps = mean(steps) )
# Plot the time series data.
hg <- ggplot(data = plot.df, aes(x = interval, y = mean.steps)) + 
  geom_line(col = 'Blue') + ggtitle('Weekday/Weekend Activity patterns') + 
  xlab('5 - Minute interval') + ylab('Average daily steps') + 
  facet_grid(day.type ~ .)
# Save the png
ggsave(filename = 'daytype pattern.png', plot = hg, device = 'png', path = 'figure/',
       width = 50, height = 20, units = 'cm', dpi = 320)
```
<img src = 'daytype pattern.png' />